Ensembler: Enabling high-throughput molecular simulations at the superfamily scale

### 2015-02-11 manuscript/ms.tex
Edited abstract.

The rapidly expanding body of available genomic and protein structural data
provides a rich resource for the field of biomolecular simulation. However,
simulations on an omics scale are not yet widely performed, partly because
software has had trouble keeping pace. For example, it should be possible to
study proteins across entire (super)families, and to do so in a way which
exploits the entire variety of available structural biology data. Here, we
present a new tool for enabling high-throughput simulation in the genomics era.
Ensembler takes any set of sequences - from a single sequence to an entire
superfamily of interest - and shepherds them through various stages of:
comparative modeling to all relevant PDB structures, reconstruction of missing
loops, addition of missing atoms, culling by close structural similarity,
assignment of protonation states, solvation, and refinement with molecular
simulation. The output is an ensemble of structures ready for subsequent
parallel or distributed molecular simulations. This automates much of the
time-consuming process of preparing protein models suitable for simulation,
while also allowing this process to be scaled to the superfamily scale.  A
particular advantage of this approach can be found in the construction of
kinetic models of conformational dynamics - such as Markov state models - for
which a diverse array of starting configurations is expected to aid sampling.
We demonstrate the power of this approach by constructing initial models for
all catalytic domains in the human tyrosine kinase family, using all kinase
catalytic domain structures from any organism as structural templates.
Ensembler should run on all major operating systems, and has been tested on
Linux and OS X. The program is free of charge, and is made available under the
terms of the GNU General Public License (GPL) v2. The latest release can be
installed via the conda package manager, and the latest source can be
downloaded from \url{https://github.com/choderalab/ensembler}.

>>>>

The rapidly expanding body of available genomic and protein structural data provides a rich resource for understanding protein dynamics with biomolecular simulation. 
However, simulations on an \emph{omics} scale are not yet widely performed, partly because software infrastructure to enable this has not kept pace. 
For example, it should now be possible to study protein dynamics across entire (super)families, exploiting the entire variety of available structural biology data and similarities in conformations that homologous proteins can adopt.
Here, we present a new tool for enabling high-throughput simulation in the genomics era.
Ensembler takes any set of sequences---from a single sequence to an entire superfamily of interest---and shepherds them through various stages of modeling and refinement to produce simulation-ready structures.
This includes comparative modeling to all relevant PDB structures (which may span multiple conformational states of interest), reconstruction of missing loops, addition of missing atoms, culling of nearly identical structures, assignment of appropriate protonation states, solvation in explicit solvent, and refinement with molecular simulation to ensure stable simulation. 
The output of this pipeline is an ensemble of structures ready for subsequent parallel or distributed molecular simulations using clusters, supercomputers, or distributed computing projects like Folding@home.
Ensembler automates much of the time-consuming process of preparing protein models suitable for simulation, while allowing scalability up to entire superfamilies.  
A particular advantage of this approach can be found in the construction of kinetic models of conformational dynamics---such as Markov state models---which benefit from a diverse array of initial configurations that span the accessible conformational states to aid sampling.
We demonstrate the power of this approach by constructing models for all catalytic domains in the human tyrosine kinase family, using all available kinase catalytic domain structures from any organism as structural templates.

Ensembler is free of charge, and is made available under the terms of the GNU General Public License (GPL) v2. 
It should run on all major operating systems, and has been tested on Linux and OS X.
The latest release can be installed via the {\tt conda} package manager, and the latest source can be downloaded from \url{https://github.com/choderalab/ensembler}.

\emph{Keywords: molecular dynamics simulation; comparative modeling}
